Numerical Evolution from the Hamiltonian

### 2020-10-05 paper/paper.tex
Added to paper

%-------------------------------------------------------------------------------
\section{Numerical procedure}
Standard numerical routines for solving systems of ODEs expect the problem to be
written in the form
\begin{equation}\label{eq:numheqs}
  \frac{d\psi_i}{dt}=G_i(t, \psi_1, \psi_2, \ldots),\quad i=1, \ldots, N,
\end{equation}
where $N$ is the number of unknown functions $\psi_i(t)$ (and the number of
equations in well posed problems) and $G_i(t, \psi_1, \psi_2, \ldots)\equiv
G_i(t,\psi)$ are the rhs of the differential equations, wich must be coded by
the user as callable subprograms. The ODE
solver routine automatically invokes these subprograms in order (as many times
as required by the sspecific method and/or desired accuracy) to compute
estimates of the unknown functions $\psi_i$.

For Hamilton's equations, the $\psi_i$ are the coordinates and momenta, while
the $G_i$ are the partial derivatives of the hamiltonian function, for which, in
the standard approach, the user must derive expressions to be programmed.
Instead, in the method now proposed, only the hamiltonian function is coded and
the ODE solver is modified so that it repeatedly invokes that subprogram in
order to compute the hamiltonian's partial derivatives numerically. Only then
does it proceed to the estimation of the ODE solution.

To make this discussion clearer, let's consider a specific problem, the
classical harmonic oscilator~\cite{french:71}, and a specific ODE solver method,
the explicit forward Euler method \cite{sir:2010}. The hamiltonian function for
a classical oscilator with mass $m$ and restoring force constant $k$ is 
\begin{equation}
  H(q,p)=\frac{1}{2}kq^2+\frac{p^2}{2m}.
\end{equation}
Hamilton's equations~\eqref{eq:heqs} then read
\begin{align}
  \frac{dq}{dt}&=\frac{p}{m}&
  \frac{dp}{dt}&=-kq,
\end{align}
wich are already in the form of eq.~\eqref{eq:numheqs}. Given initial values
$q_0$ and $p_0,$ successive values can be estimated at times $t_1$, $t_2$, \ldots
by iterating the formulas
\begin{align}
  q_{i+1}&=q_{i}+\frac{p_i}{m}\delta t_i\label{eq:efe1}\\
  p_{i+1}&=p_{i}-kq_i\delta t_i\label{eq:efe2},
\end{align}
where $\delta t_i=t_{i+1}-t_i$. It is now easy to write a  program that
implements this method, and this the usual procedure. With this paper, we
consider a method where explicit expressions for the hamiltonian's partial
derivatives in the equations of motion are replaced by numerical estimates. In
our toy problem, that amounts to rewriting eqs~\eqref{eq:efe1}
and~\eqref{eq:efe2} by
\begin{align*}
  q_{i+1}&=q_{i}+\numap{\frac{\partial H}{\partial p}}_i \delta t_i\\
  p_{i+1}&=p_{i}-\numap{\frac{\partial H}{\partial q}}_i\delta t_i,
\end{align*}
where $\numap{X}_i$ means \emph{numerical approximation of $X$ at time $t_i$.}
In order to make numerical aproximations to the partial derivatives of a
function, one must evaluate it several times; at the very least, two evaluations
are needed. For example, using the central difference method, the derivative is
estimated as
\begin{equation} 
  \frac{\partial\psi}{\partial x} = 
    \frac{\psi(x+\delta x)-\psi(x-\delta t)}{2\delta x}+
  o(\delta x)^2,
\end{equation}
leading to approximation formulas that, for our example problem, read 
\begin{align}
  \numap{\frac{\partial H}{\partial p}}_i &=
  \frac{H(q_i, p_i+\delta p)-H(q_i, p_i-\delta p)}{2\delta p}\\
  \numap{\frac{\partial H}{\partial q}}_i &=
  \frac{H(q_i+\delta q, p_i)-H(q_i-\delta q, p_i)}{2\delta q},
\end{align}
where $\delta q$ and $\delta p$ are suitably chosen discretization parameters.
Other approximations have higher accuracy, at the cost of extra evaluations of
the function to be derived.

@@

\section{Simple implementation in \texttt{python} and examples}
\section{Conclusion}

\section*{References}
\begin{thebibliography}{99}
  \bibitem{gol:1980} H. Goldstein, \textsl{Classical Mechanics.} Addison-Wesley
  (1980)
  \bibitem{french:71} A.P.~French, \textsl{Vibration and Waves.} W.W. Norton \&
  Company (1971)
  \bibitem{sir:2010} S. \v{S}irca, M. Horvat, \textsl{Computational Methods in
  Physics.} Springer (2012)
  \bibitem{nr:2007} W.H.~Press, S.A.~Teukolsky, W.T.~Vetterling, B.P.~Flannery,
  \textsl{Numerical Recipes -- The Art of Scientific Computing,} 3rd Ed.
  Cambridge University Press (2007)
\end{thebibliography}
\end{document}

>>>>

%-------------------------------------------------------------------------------
\section{Numerical procedure}
Standard numerical routines for solving systems of ODEs expect the problem to be
written in the form
\begin{equation}\label{eq:numheqs}
  \frac{d\psi_i}{dt}=G_i(t, \psi_1, \psi_2, \ldots),\quad i=1, \ldots, N,
\end{equation}
where $N$ is the number of unknown functions $\psi_i(t)$ (and the number of
equations in well posed problems) and $G_i(t, \psi_1, \psi_2, \ldots)\equiv
G_i(t,\psi)$ are the rhs of the differential equations, wich must be coded by
the user as callable subprograms. During its execution, the ODE solver routine
repeatedly invokes these subprograms (as many times as required by the specific
integration algorithm and/or desired accuracy) in order to compute estimates for
the values the unknown functions $\psi_i$ at given values of $t$.

For Hamilton's equations, the $\psi_i$ are the coordinates and conjugate momenta
of the system, while the $G_i$ are the partial derivatives of its hamiltonian
function. The usual procedure is to derive analitical expressions for the $G_i$
as functions of $t$, the coordinates and the momenta, and then to code those
expressions as subprograms to be called by the ODE solver routine.

In the method now proposed, instead, the user only has to supply code to
compute the value of the system's hamiltonian function; its partial derivatives
(the $G_i$ in eq.~\eqref{eq:numheqs}) are estimated numerically by another
subprogram called by the ODE solver.

%is coded and
%the ODE solver is modified so that it repeatedly invokes that subprogram in
%order to compute the hamiltonian's partial derivatives numerically. Only then
%does it proceed to the estimation of the ODE solution.

To make this discussion clearer, let us consider a specific problem, the
classical harmonic oscilator~\cite{french:71}, and a specific ODE solver method,
the explicit forward Euler method \cite{sir:2010}. The hamiltonian function for
a classical oscilator with mass $m$ and restoring force constant $k$ is 
\begin{equation}
  H(q,p)=\frac{1}{2}kq^2+\frac{p^2}{2m}.
\end{equation}
Hamilton's equations~\eqref{eq:heqs} then read
\begin{align}
  \frac{dq}{dt}&=\frac{p}{m}&
  \frac{dp}{dt}&=-kq,
\end{align}
wich are already in the form of eq.~\eqref{eq:numheqs}. Given initial values
$q_0$ and $p_0,$ successive values can be estimated at times $t_1$, $t_2$, \ldots
by iterating the formulas
\begin{align}
  q_{i+1}&=q_{i}+\frac{p_i}{m}\delta t_i\label{eq:efe1}\\
  p_{i+1}&=p_{i}-kq_i\delta t_i\label{eq:efe2},
\end{align}
where $\delta t_i=t_{i+1}-t_i$. It is now easy to write a  program that
implements this method, and this the usual procedure. With this paper, we
consider a method where explicit expressions for the hamiltonian's partial
derivatives in the equations of motion are replaced by numerical estimates. In
our toy problem, that amounts to rewriting eqs~\eqref{eq:efe1}
and~\eqref{eq:efe2} by
\begin{align*}
  q_{i+1}&=q_{i}+\numap{\frac{\partial H}{\partial p}}_i \delta t_i\\
  p_{i+1}&=p_{i}-\numap{\frac{\partial H}{\partial q}}_i\delta t_i,
\end{align*}
where $\numap{X}_i$ means \emph{numerical approximation of $X$ at time $t_i$.}
In order to make numerical aproximations to the partial derivatives of a
function, one must evaluate it several times; at the very least, two evaluations
are needed. For example, using the second order central difference method, the
derivative is estimated as
\begin{equation}\label{eq:cdiff}
  \frac{\partial\psi}{\partial x} = 
    \frac{\psi(x+\delta x)-\psi(x-\delta t)}{2\delta x}+
  o(\delta x)^2,
\end{equation}
leading to approximation formulas that, for our example problem, read 
\begin{align}
  \numap{\frac{\partial H}{\partial p}}_i &=
  \frac{H(q_i, p_i+\delta p)-H(q_i, p_i-\delta p)}{2\delta p}\\
  \numap{\frac{\partial H}{\partial q}}_i &=
  \frac{H(q_i+\delta q, p_i)-H(q_i-\delta q, p_i)}{2\delta q},
\end{align}
where $\delta q$ and $\delta p$ are suitably chosen discretization parameters.
Other approximations have higher accuracy, at the cost of extra evaluations of
the function to be derived.

@@

\section{Simple implementation in Python and examples}
We made a simple implementation of this method in Python, using the standard
numerical extensions library NumPy~\cite{numpy:2011,numpy:2020}.  It consists of
a simple function that calls the hamiltonian function supplied by the user to
evaluate its partial derivatives, using the central difference formulas of
eqs.~\eqref{eq:cdiff}. This function is then supplied to NumPy's ODE solver
\texttt{solve\_ivp}.  For greater efficiency, it is wrapped in a class, whose
objects store initialization details on creation, like the hamiltonian's
parameters, the number of degrees of freedom, the values of the discretization
steps, etc. These objects are callable with signature adjusted to what the ODE
solver expects.  The whole class definition fits in a few dozen lines of code
and is very easy to use. It is available at

\section{Conclusion}

\section*{References}
\begin{thebibliography}{99}
  \bibitem{gol:1980} H. Goldstein, \textsl{Classical Mechanics.} Addison-Wesley
  (1980)
%
  \bibitem{french:71} A.P.~French, \textsl{Vibration and Waves.} W.W. Norton \&
  Company (1971)
%
  \bibitem{sir:2010} S. \v{S}irca, M. Horvat, \textsl{Computational Methods in
  Physics.} Springer (2012)
%
  \bibitem{nr:2007} W.H.~Press, S.A.~Teukolsky, W.T.~Vetterling, B.P.~Flannery,
  \textsl{Numerical Recipes -- The Art of Scientific Computing,} 3rd Ed.
  Cambridge University Press (2007)
%
  \bibitem{numpy:2011} S.J.~van der Walt, S.C.~Colbert, G.~Varoquaux,
  \textsl{The NumPy array: a structure for efficient numerical computation,}
  Computing in Science \& Engineering, \textbf{13} no. 2, 22-30 (2011).
  doi: 10.1109/MCSE.2011.37
%
  \bibitem{numpy:2020} C.R.~Harris, K.J.~Millman, S.J.~van der Walt et. al.,
  \textsl{Array programming with NumPy,} Nature \textbf{585}, 357-362 (2020).
  https://doi.org/10.1038/s41586-020-2649-2
\end{thebibliography}
\end{document}
